---
title: "2. DMM: Dirichlet-Multinomial Mixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. DMM: Dirichlet-Multinomial Mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

The Dirichlet-Multinomial Mixture answers **how many discrete community types
the dataset supports**, via formal model selection. It is a *hard-assignment*
model: each sample belongs to one component. This vignette covers every
DMM-specific function; generalist plots (`plot_pcoa`, `plot_barplot`, ...) are
shown in `vignette("generalist")`.

```{r}
library(symbayes)
sp <- import("run.seqs.absolute.abund_only.txt",
             profs_abund = "run.profiles.absolute.abund_only.txt",
             profs_meta  = "run.profiles.meta_only.txt")
sp <- filter_samples(sp)
sp <- fit_dmm(sp, max_k = 10)
```

## Model selection

`fit_dmm()` evaluates `k = 1:max_k` and selects K by the Laplace approximation
to the model evidence. Inspect the criteria:

```{r}
dmm_plot_selection(sp)
sp$dmm$meta$model_fit     # the Laplace / AIC / BIC table
```

## Assignment and certainty

The membership object holds the posterior and per-sample certainty:

```{r}
sp$dmm$k                   # selected number of components
head(sp$dmm$theta)         # posterior membership (near one-hot)
head(sp$dmm$meta$certainty)
```

DMM certainty is typically near 1.0 even for mixed samples: this is a property
of the likelihood at amplicon read depths, not evidence of unmixed communities.
The tools below probe whether that confidence is *stable*.

## Prediction on new samples

Score new count vectors against the fitted components with no refitting
(closed-form Dirichlet-Multinomial predictive):

```{r}
# A hypothetical C22-dominated sample
model_seqs <- colnames(sp$count_mat)
x <- setNames(rep(0L, length(model_seqs)), model_seqs)
x[c("C22", "C3", "C1")] <- c(3000, 1500, 500)

predict_dmm(sp, x)
```

## Leave-one-out cross-validation

Refit on N-1 samples, predict the held-out one; reports concordance and flags
samples that switch component (candidate boundary/mixed samples):

```{r}
loo <- dmm_loo(sp)
subset(loo, !concordant)
```

## Subsample cross-validation

Faster alternative: repeatedly hold out a fraction and predict.

```{r}
stab <- dmm_subsample_cv(sp, frac_test = 0.2, n_reps = 50)
head(stab)                 # least stable samples first
```

## Simulation

Three modes probe DMM behaviour.

**Blend** two components across a gradient to trace the decision boundary:

```{r}
blend <- dmm_simulate(sp, mode = "blend", blend_comps = c(1, 2),
                      blend_range = seq(0, 1, 0.05))
```

**Perturb** real samples (reassign a fraction of reads) to test robustness — a
predictively-neutral perturbation that flips many "certain" assignments
demonstrates hard-assignment fragility:

```{r}
perturb <- dmm_simulate(sp, mode = "perturb", perturb_frac = 0.10)
mean(perturb$assigned == perturb$true_comp)    # accuracy under 10% noise
```

**From alpha** draws from each component's own Dirichlet — a self-consistency
check:

```{r}
dmm_simulate(sp, mode = "from_alpha", n_sim = 200)
```
