---
title: "4. HDP: Hierarchical Dirichlet Process"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. HDP: Hierarchical Dirichlet Process}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

The HDP is the threshold-free, mixture-aware core of `symbayes`. The number of
community types is **inferred from the data**, not specified, and samples may
be admixed. It requires the `hdp` package:

```{r}
# remotes::install_github("nicolaroberts/hdp")
library(symbayes)
sp <- import("run.seqs.absolute.abund_only.txt",
             profs_abund = "run.profiles.absolute.abund_only.txt",
             profs_meta  = "run.profiles.meta_only.txt")
sp <- filter_samples(sp)
```

## Fitting

By default the structure is **flat** (one parent DP, one child DP per sample):
assignment is covariate-free, which is the right choice for host-blind claims.

```{r}
sp <- fit_hdp(sp)             # flat, threshold-free
sp$hdp$k                       # inferred number of components
```

### Grouped structure (use with care)

Supplying `group_var` inserts a middle DP layer so samples within a group share
strength. This improves prediction but introduces **circularity** if you later
test that same covariate's effect on community. Use flat for core claims;
grouped only as an explicitly-labelled predictive enhancement.

```{r}
sp_grouped <- fit_hdp(sp, group_var = "host_species")
```

## Inferred component number

The key threshold-free evidence: the sampler explores a distribution of cluster
numbers; extraction merges cosine-similar and drops transient clusters to yield
the stable components. The raw-to-extracted reduction is expected.

```{r}
hdp_plot_numcomp(sp)
sp$hdp$meta$raw_ncl            # sampler's median raw cluster count
```

If the extracted count is far below the raw median, extraction may be merging
too hard; raise `cos_merge` toward 0.95 and refit.

## Fractional membership and residual

```{r}
head(sp$hdp$theta)             # fractional; residual held separately
head(sp$metadata[, c("hdp_dominant", "hdp_entropy",
                     "hdp_max_prop", "hdp_n_comp", "hdp_residual")])
```

`hdp_residual` is the mass a sample places on *no* component — honest
uncertainty SymPortal cannot express.

## Over-splitting quantification

The core method metric. It partitions SymPortal profiles into three mutually
exclusive classes so they don't contaminate one another: confident
over-splits, mixtures, and orphans.

```{r}
res <- hdp_vs_symportal(sp, over_split_min_frac = 0.7, mixture_thresh = 0.7)

res$oversplit_ratio            # SymPortal profiles / HDP components
res$redundant_splits           # HDP components that absorb >1 confident profile
res$mixture_profiles           # profiles that are HDP admixtures
res$orphan_profiles            # rare/poorly-fit profiles (excluded from count)
```

Report the mixture claim on reproducible profiles (n >= 2); fold n = 1 profiles
into the orphan/uncertain category.

## Corroborating with LDA

Two threshold-free-oriented methods agreeing on K is strong evidence. Compare
HDP's inferred component number to LDA's redundancy-selected K:

```{r}
sp <- fit_lda(sp)
compare_models(sp)
```
