---
title: "3. LDA: Latent Dirichlet Allocation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. LDA: Latent Dirichlet Allocation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

LDA answers **what each sample is made of**. Unlike DMM, membership is
*fractional*, so within-sample admixture is represented directly. This vignette
covers every LDA-specific function.

```{r}
library(symbayes)
sp <- import("run.seqs.absolute.abund_only.txt",
             profs_abund = "run.profiles.absolute.abund_only.txt",
             profs_meta  = "run.profiles.meta_only.txt")
sp <- filter_samples(sp)
```

## Adaptive topic-number selection

With `k = NULL` (the default), `fit_lda()` searches `k_range` and selects the
**largest K with no redundant topic pairs** (cosine `< redundant_thresh`).
This resists the shadow topics that perplexity-optimal K produces.

```{r}
sp <- fit_lda(sp, k = NULL, k_range = 2:15, redundant_thresh = 0.85)
lda_plot_ksearch(sp)          # 4-panel K diagnostics
sp$lda$meta$selection         # the search table
```

To fix K instead:

```{r}
sp <- fit_lda(sp, k = 9)
```

## Topic redundancy

The cosine-similarity matrix among topic exemplars; pairs above ~0.85 signal
over-splitting. `fit_lda()` stores it and the flagged pairs.

```{r}
lda_plot_similarity(sp)
sp$lda$meta$redundant         # flagged redundant pairs (should be empty)
```

## Fractional membership

```{r}
sp$lda$k
head(sp$lda$theta)            # rows sum to 1, values fractional
head(sp$metadata[, c("lda_dominant", "lda_entropy",
                     "lda_max_prop", "lda_n_topics")])
```


## Topic summary

A text report cross-referencing each topic to clade, host species, SymPortal
profiles, and its exemplar composition:

```{r}
lda_topic_summary(sp, top_n = 8)
```

## Interpreting exemplar overlap

Two topics sharing a sequence backbone are not necessarily redundant. Use the
cosine matrix (redundancy), per-sample `theta` (mixing), and pure-sample counts
together — a multi-coloured exemplar can reflect real intragenomic variation
within one community type rather than mixing. See the generalist
`plot_exemplars(sp, model = "lda")` and `plot_membership(sp, model = "lda")`.
